---
title: "p8105_hw2_xc2859"
author: "Xiaoxue Cui"
date: "2026-10-05"
output: github_document
---

# Problem 1
__Read and clean the data__

```{r}
library(tidyverse)

nycsubway_df = 
  read.csv(
    "data/NYC_Transit_Subway_Entrance_And_Exit_Data.csv", 
    na = c("NA", ".", "")) |>
  janitor::clean_names() |>
  select(line, station_name, station_latitude, station_longitude, route1, route2, route3, route4, route5, route6, route7, route8, route9, route10, route11, entrance_type, entry, vending, ada) |>
  mutate(
    entry = case_match(
      entry, 
      "YES" ~ TRUE, 
      "NO" ~ FALSE
    )
  )
```
The dataset contains information on subway lines, stations, and station entrances. Variables include `line`, `station_name`, `station_latitude`, `station_longitude`, routes served (`route1` to `route11`), `entry` (whether the entrance allows entry into the station), `vending` (vending machine availability), `entrance_type`, as well as `ada` (ADA compliance). For data cleaning, I first cleaned the variable names by using `clean_names` function, then selected the variables needed for the analysis by using `select` function, and finally converting `entry` from a character variable to a logical variable by applying `mutate` and `case_match` functions. The resulting dataset has `r nrow(nycsubway_df)` rows and `r ncol(nycsubway_df)` columns. These data are not tidy because information on routes served are stored in multiple columns instead of in separate route number and route name variables. 

__Answer the questions__

```{r}
# How many distinct stations are there?
nycsubway_df |> 
  distinct(line, station_name) |> 
  nrow()

# How many stations are ADA compliant?
nycsubway_df |>
  filter(ada == TRUE) |>
  distinct(line, station_name) |>
  nrow()

# What proportion of station entrances / exits without vending allow entrance?
nycsubway_df |>
  filter(vending == "NO") |>
  pull(entry) |> 
  mean()
```
Therefore, there are 465 distinct stations. 84 stations are ADA compliant. There are about 37.70% station entrances / exits without vending allow entrance.  


__Reformat data__
```{r}

```

