---
title: "p8105_hw2_xc2859"
author: "Xiaoxue Cui"
date: "2026-10-05"
output: github_document
---

# Problem 1
__Read and clean the data__

```{r}
library(tidyverse)

nycsubway_df = 
  read.csv(
    "data/NYC_Transit_Subway_Entrance_And_Exit_Data.csv", 
    na = c("NA", ".", "")) |>
  janitor::clean_names() |>
  select(line, station_name, station_latitude, station_longitude, route1, route2, route3, route4, route5, route6, route7, route8, route9, route10, route11, entrance_type, entry, vending, ada) |>
  mutate(
    entry = case_match(
      entry, 
      "YES" ~ TRUE, 
      "NO" ~ FALSE
    )
  )
```
The dataset contains information on subway lines, stations, and station entrances. Variables include `line`, `station_name`, `station_latitude`, `station_longitude`, routes served (`route1` to `route11`), `entry` (whether the entrance allows entry into the station), `vending` (vending machine availability), `entrance_type`, as well as `ada` (ADA compliance). For data cleaning, I first cleaned the variable names by using `clean_names` function, then selected the variables needed for the analysis by using `select` function, and finally converting `entry` from a character variable to a logical variable by applying `mutate` and `case_match` functions. The resulting dataset has `r nrow(nycsubway_df)` rows and `r ncol(nycsubway_df)` columns. These data are not tidy because information on routes served are stored in multiple columns instead of in separate route number and route name variables. 

__Answer the questions__

```{r}
# How many distinct stations are there?
nycsubway_df |> 
  distinct(line, station_name) |> 
  nrow()

# How many stations are ADA compliant?
nycsubway_df |>
  filter(ada == TRUE) |>
  distinct(line, station_name) |>
  nrow()

# What proportion of station entrances / exits without vending allow entrance?
nycsubway_df |>
  filter(vending == "NO") |>
  pull(entry) |> 
  mean()
```
Therefore, there are 465 distinct stations. 84 stations are ADA compliant. There are about 37.70% station entrances / exits without vending allow entrance.  


__Reformat data__
```{r}
nycsubway_df = 
  nycsubway_df |> 
  mutate(
    across(route1:route11, as.character)
  )

nycsubway_tidy_df = 
  nycsubway_df |>
  pivot_longer(
    route1:route11, 
    names_to = "route_number", 
    names_prefix = "route", 
    values_to = "route_name", 
    values_drop_na = TRUE
  )

# Number of distinct stations that serve the A train
nycsubway_tidy_df |>
  filter(route_name == "A") |>
  distinct(line, station_name) |> 
  nrow()

# Of the stations that serve the A train, how many are ADA compliant?
nycsubway_tidy_df |>
  filter(
    route_name == "A", 
    ada == TRUE) |>
  distinct(line, station_name) |> 
  nrow()
```
There are 60 distinct stations that serve the A train. Of the stations that serve the A train, 17 are ADA compliant. 


# Problem 2
__Create a tidy dataset__
```{r}
# Read and clean the Mr.Trash Wheel sheet
MrTrash_df = 
  readxl::read_excel(
    "data/202509 Trash Wheel Collection Data.xlsx", 
    sheet = "Mr. Trash Wheel", 
    range = "A2:N709") |> 
  janitor::clean_names() |>
  filter(!is.na(dumpster)) |>
  mutate(
    sports_balls = 
      as.integer(round(sports_balls)), 
    trash_wheel_name = "Mr. Trash Wheel", 
    year = as.integer(year)
  )

# Read and clean the Professor Trash Wheel sheet
ProfessorTrash_df = 
  readxl::read_excel(
    "data/202509 Trash Wheel Collection Data.xlsx", 
    sheet = "Professor Trash Wheel", 
    range = "A2:M135") |>
  janitor::clean_names() |> 
  filter(!is.na(dumpster)) |>
  mutate(
    trash_wheel_name = "Professor Trash Wheel", 
    year = as.integer(year)
  )

# Read and clean the Gwynnda sheet
Gwynns_df = 
   readxl::read_excel(
    "data/202509 Trash Wheel Collection Data.xlsx", 
    sheet = "Gwynns Falls Trash Wheel", 
    range = "A2:L351") |>
  janitor::clean_names() |> 
  filter(!is.na(dumpster)) |> 
  mutate(
    trash_wheel_name = "Gwynns Falls", 
    year = as.integer(year)
  )

# Combine these sheets into a tidy dataset
trashwheel_tidy_df = 
  bind_rows(MrTrash_df, ProfessorTrash_df, Gwynns_df) |>
  select(trash_wheel_name, everything())
```

__Data description__

The combined Trash Wheel dataset contains `r nrow(trashwheel_tidy_df)` observations and `r ncol(trashwheel_tidy_df)` variables. These variables include the Trash Wheel name, dumpster number, collection date, trash weight and volume, counts of different types of trash such as polystyrene waste, plastic bottles, cigarette butts, glass bottles and plastic bags, as well as estimated number of homes that could be powered by the collected trash. 

```{r}
# total weight of trash collected by Professor Trash Wheel
trashwheel_tidy_df |>
  filter(trash_wheel_name == "Professor Trash Wheel") |> 
  pull(weight_tons) |>
  sum(na.rm = TRUE)

# total number of cigarette butts collected by Gwynnda in June of 2022
trashwheel_tidy_df |> 
  filter(
    trash_wheel_name == "Gwynns Falls", 
    month == "June", 
    year == 2022
  ) |> 
  pull(cigarette_butts) |> 
  sum(na.rm = TRUE)
```
For available data, the total weight of trash collected by Professor Trash Wheel was 282.26 tons. The total number of cigarette butts collected by Gwynnda in June of 2022 was 18120. 


# Problem 3
__Import, clean, and tidy the baseline demographics dataset__
```{r}
baseline_df = 
  read.csv(
    "data_mci/MCI_baseline.csv", 
    skip = 1, 
    na = c(".", "NA", "")) |>
  janitor::clean_names() |>
  filter(
    is.na(age_at_onset) | age_at_onset > current_age
  ) |>
  mutate(
    sex = case_match(
      sex, 
      0 ~ "female", 
      1 ~ "male"
    ), 
    apoe4 = case_match(
      apoe4, 
      0 ~ "non-carrier", 
      1 ~ "carrier"
    )
  )
```

__Describe the import process and dataset features__

The baseline demographics dataset was imported using `read.csv`, with the first row skipped because it contains notes on how the variable was coded rather than meaningful data, `"."`, `"NA"` and blank entries were treated as missing values during import. Variable names were cleaned using `clean_names`. Participants who had MCI onset at baseline were excluded by retaining those who either had no recorded age at MCI onset or developed MCI after their current age. The numeric codes for variables `sex` and `apoe4` were recoded as `female`/ `male` and `non-carrier` / `carrier`, respectively. The resulting dataset has `r nrow(baseline_df)` observations and `r ncol(baseline_df)` variables. These variables contains information about participants' demographics at baseline level, including participant ID, current age, sex, years of education, APOE4 carrier status, and age at MCI onset (missing data for this variable represents that a subject remains MCI-free during the follow-up period). 

__Baselne characteristics__
```{r}
# Among recruited participants, the number of those develop MCI
baseline_df |>
  filter(
    !is.na(age_at_onset)) |>
  nrow()

# Average baseline age
baseline_df |> 
  pull(current_age) |>
  mean() |> 
  round(digits = 2)

# Proportion of women who are APOE4 carriers
baseline_df |> 
  filter(sex == "female") |>
  mutate(
    apoe4 = case_match(
      apoe4, 
      "carrier" ~ TRUE, 
      "non-carrier" ~ FALSE
    )
  ) |>
  pull(apoe4) |>
  mean() 
```
There were `r nrow(baseline_df)` participants recruited. Of those recruited, 93 participants developed MCI. The average baseline age is 65.03. 30% of women in the study are APOE4 carriers. 

__Import, clean, and tidy the dataset of longitudinally observed biomarker values__
```{r}
longitudinal_df = 
  read.csv(
    "data_mci/mci_amyloid.csv", 
    skip = 1, 
    na = c(".", "NA", "")) |>
  janitor::clean_names() |> 
  pivot_longer(
    baseline:time_8, 
    names_to = "time_years", 
    values_to = "value"
  ) |> 
  mutate(
    time_years = recode_values(
      time_years, 
      "baseline" ~ 0, 
      "time_2" ~ 2, 
      "time_4" ~ 4, 
      "time_6" ~ 6, 
      "time_8" ~ 8
    )
  )
```

__Describe the import process and dataset features__

The longitudinal amyloid value dataset was imported using `read.csv`, with the first row skipped since it did not contain information about the data. `"."`, `"NA"` and blank entries were treated as missing values during import, and variable names were cleaned using `clean_names`. The repeated biomarker measurements, originally stored in separate columns for baseline and follow-up visits, were reformatted using `pivot_longer`. The visit variable was then recoded as `time_years`, with 0 representing baseline and 2, 4, 6, and 8 representing years elapsed since the study baseline. The resulting dataset is now tidy. Key variables include participant ID, time in years passed since the baseline time, and the corresponding results of amyloid β 42/40 ratio measurement (called `values`). 